---
title: "Illapani_IS606_Proposal"
author: "Srini Illapani"
date: "October 16, 2015"
output: pdf_document
---



```{r include=FALSE, cache=FALSE}
# DO NOT REMOVE
# THIS IS FOR SETTING SOME PLOTTING PARAMETERS SO THAT YOUR PLOTS DON'T TAKE UP TOO MUCH SPACE
# IF YOU WOULD LIKE TO CHANGE THESE, SEE HELP FILES ON THE par() FUNCTION
# OR ASK FOR HELP
library(knitr)
## set global chunk options
opts_chunk$set(fig.path='figure/manual-', cache.path='cache/manual-', fig.align='center', fig.show='hold', par=TRUE)
## tune details of base graphics (http://yihui.name/knitr/hooks)
knit_hooks$set(par=function(before, options, envir){
if (before && options$fig.show!='none') par(mar=c(4,4,.2,.1),cex.lab=.95,cex.axis=.9,mgp=c(2,.7,0),tcl=-.3)
})
```

```{r, eval=TRUE}
# Load the necessary packages to be used for this project
library(dplyr)
library(tidyr)
library(knitr)
library(ggplot2)
```



```{r, eval=TRUE}
# load data - Upload the file to my github repository and use the curl package to retrieve the dataset in to R.
library(curl)

endorsements <- read.csv(curl("https://raw.githubusercontent.com/isrini/SI_IS606/master/endorsements.csv"), header = TRUE)

kable(head(endorsements,10))
```



### Research question 

  Do number of endorsements increase the chances of winning the primaries and party nomination?  
  We would also attempt to see if the money raised by the candidates is any indicator of winning the primaries, if time permits.  


### Cases 

**What are the cases, and how many are there?**  

  We will be looking at the democratic and republican party candidates for presidential primaries from
  the year 1980 to 2012.



### Data collection 

**Describe the method of data collection.**

  Fundraising data since 1992 is taken from the Federal Election Commission website. Data from 1980 to 1988
  is from various news    articles at the time of the filing deadline. This data is preliminary, though rarely
  differs greatly from finalized data.  
  
  All the data used excludes self-funding by the candidates.
  
  

### Type of study 

**What type of study is this (observational/experiment)?**  

  The presidential candidate endorsements for primary elections is an observational study performed by
  FiveThirtyEight by collecting the data as mentioned above.


### Data Source  

  The data souce for the puropose of this Data project is FiveThirtyEight (https://github.com/fivethirtyeight/data).
  Here is the list of variables that will be used for the project:  
  
  
Value | Description
------|------------
year|Election year
party|Political party
candidate|Candidate running in primary
endorsement_points|Weighted endorsements through June 30th of the year before the primary
percentage_endorsement_points|Percentage of total weighted endorsement points for the candidate's political party through June 30th of the year before the primary
money_raised|Money raised through June 30th of the year before the primary
percentage_of_money|Percentage of total money raised by the candidate's political party through June 30th of the year before the primary
primary_vote_percentage|Percentage of votes won in the primary
won_primary|Did the candidate win the primary?



### Response 

**What is the response variable, and what type is it (numerical/categorical)?**

  The response variable for the question we are trying to answer - Do number of endorsements increase the chances of winning the    primaries? is 'won_primary'.  
  
  It is of type categorial, and can also be treated as numerical by converting to 0 or 1.


### Explanatory 

**What is the explanatory variable, and what type is it (numerical/categorical)?**  

  The explanatory variables in the data set are 'endorsement_points' and 'money_raised'.
  These are of type numerical.  

### Relevant summary statistics 

**Provide summary statistics relevant to your research question. For example, if you're comparing means across groups provide means, SDs, sample sizes of each group. This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r}
# Sample statistics to be run answer the research question:

# Using group by and summarise to find the mean endorsements by the year and party
mean = group_by(endorsements,year) %>% 
    filter(endorsement_points != 0) %>% 
    summarise(mean(endorsement_points))

head(mean)
```

